Provably Correct, Low-Cost Optimizations for Abstract Abstract Machines

### 2013-03-25 paper/paper.tex
Added more of a teaser to the end of the intro.

However, there is a considerable weakness with the approach: an
analyzer designed and implemented by following the AAM recipe is
prohibitively inefficient without both further approximation and
further implementation effort.

In this article, we develop a systematic approach to deriving a
feasible implementation of an abstract-machine-based analyzer.

@@

At each step during the initial presentation and development, we
evaluated the implementation on a set of benchmarks. The highlighted
benchmark in figure \ref{fig:churchtime} is from Vardoulakis and
Shivers~\cite{dvanhorn:Vardoulakis2011CFA2} that tests distributivity
of multiplication over addition on Church numerals.  For the
step-by-step development, this benchmark is particularly informative:
\begin{enumerate}
\item it can be written in most modern programming languages,
%
\item it was designed to stress an analyzer's ability to deal with
  complicated environment and control structure arising from the use
  of higher-order functions to encode arithmetic, and
%
\item its improvement is about median in the benchmark suite
  considered in section~\ref{sec:eval}, and thus it serves as a good
  sanity check for each of the optimization techniques considered.
\end{enumerate}

@@

Conceptually, the AAM approach calls for computing an analysis as a
graph exploration: (1) start with an initial state, and (2) compute
the transitive closure of the transition relation from that state. All
visited states are potentially reachable in the concrete, and all
paths through the graph are possible traces of execution.

We can cast this exploration process in terms of a fixed-point calculation.
%
Given the initial state $\mstate_0$ and the transition relation $\machstep$,
we define the global transfer function:
\begin{equation*}
 F_{\mstate_0} : \mathcal{P}(\State) \to \mathcal{P}(\State)
 \text.
\end{equation*}
Internally, this global transfer function computes the successors of all supplied states, and then includes the initial state:
\begin{align*}
  F_{\mstate_0}(V,E) &= (\{ \mstate_0 \} \cup V', E') \\
    E' &= \setof{ (\mstate,\mstate') \mid \mstate \in V \text{ and } \mstate \machstep \mstate'} \\
    V' &= \setof{ \mstate' \mid (\mstate,\mstate') \in E'}
\end{align*}
Then, the evaluator for the analysis computes the least fixed-point of the global transfer function:
\begin{equation*}
 \eval(\mexp) = \mathrm{lfp}(F_{\mstate_0})\text{,}
\end{equation*}
where $\mstate_0 = \ev[^\mtcntr]{\mexp, \varnothing, \varnothing, \kmt}$.

The possible traces of execution tell us the most about a program, so
we take $\traces(\mexp)$ to be the (regular) set of paths through the
computed graph.

@@

A common technique to accelerate convergence in flow analyses is to share a
common, global store.
%
To retain soundness, this store grows monotonically.
%
Formally, we can cast this optimization as a second abstraction or as the
application of a widening operator during the fixed-point iteration.
%
In the ISWIM language, such a widening makes 0-CFA quartic in the size of the
program.
%
Thus, in one step, complexity drops from intractable exponentiality to a merely
daunting polynomial.

@@

The only analyzers we were able to find that proved capable of
analyzing the full suite of benchmarks considered here were the Soft
Typing system of Wright and
Cartwright~\cite{dvanhorn:Wright1997Practical} and, in many ways its
successor, the Polymorphic splitting system of Wright and
Jagannathan~\cite{dvanhorn:wright-jagannathan-toplas98}.\footnote{This
  is not a coincidence; these papers set a high standard for
  evaluation, which we consciously aimed to approach.}  Unfortunately,
these analyses compute an inherently different and incomparable form
of analysis.  Consequently, we have omitted a complete comparison with
these implementations.  The AAM approach provides more precision in
terms of temporal-ordering of program states, which comes at a cost
that can be avoided in constraint-based approaches.  Consequently
implementation techniques cannot be ``ported'' between these two
approaches.  However, our optimized implementation is within an order
of magnitude of the performance of Wright and Jaganathan's analyzer.
Although we would like to improve this to be more competitive, the
optimized AAM approach still has many strengths to recommend it in
terms of precision, ease of implementation and verification, and rapid
design. We can get closer to their performance by relying on the
representation of addresses and the behavior of $\alloc$ to
pre-allocate most data structures and split the abstract store out
into parts that are more quickly accessed and updated. The first two
of our optimizations can still be applied to an analysis that does
abstract garbage collection~\cite{dvanhorn:Might:2006:GammaCFA},
whereas the polymorphic splitting implementation is tied strongly to a
single-threaded store.

@@

Boucher and Feeley \cite{dvanhorn:Boucher1996Abstract} introduced the
idea of abstract compilation, which used closure generation
\cite{dvanhorn:Feeley1987Using} to improve the performance of control
flow analysis.  We have adapted the closure generation technique from
compositional evaluators to abstract machines and applied it to similar
effect.

>>>>

However, there is a considerable weakness to the approach: an
analyzer designed and implemented by following the AAM recipe is
prohibitively inefficient without both further approximation and
further implementation effort.

In this article, we develop a systematic approach to deriving a
practical implementation of an abstract-machine-based analyzer using
mostly semantic means - not tricky engineering. Our goal is to empower
programming language implementers and researchers to explore and
convincingly exhibit their ideas with a low barrier to entry. The
optimizations we describe are widely applicable and apparently
effective to scale far beyond toy examples.

@@

At each step during the initial presentation and development, we
evaluated the implementation on a set of benchmarks. The highlighted
benchmark in figure \ref{fig:churchtime} is from Vardoulakis and
Shivers~\cite{dvanhorn:Vardoulakis2011CFA2} that tests distributivity
of multiplication over addition on Church numerals.  For the
step-by-step development, this benchmark is particularly informative:
\begin{enumerate}
\item it can be written in most modern programming languages,
%
\item it was designed to stress an analyzer's ability to deal with
  complicated environment and control structure arising from the use
  of higher-order functions to encode arithmetic, and
%
\item it proves to be the \emph{least} improved benchmark of the
  complete suite considered in section~\ref{sec:eval}, and thus it
  serves as a good sanity check and lower-bound for each of the
  optimization techniques considered.
\end{enumerate}

@@

Conceptually, the AAM approach calls for computing an analysis as a graph
exploration: (1) start with an initial state, and (2) compute the transitive
closure of the transition relation from that state.

We can cast this exploration process in terms of a fixed-point calculation.
%
Given the initial state $\varsigma_0$ and the transition relation $\machstep$,
we define the global transfer function:
\begin{equation*}
 F_{\varsigma_0} : \mathcal{P}(\State) \to \mathcal{P}(\State)
 \text.
\end{equation*}
Internally, this global transfer function computes the successors of all supplied states, and then includes the initial state:
\begin{equation*}
  F_{\varsigma_0}(S) = \{ \varsigma_0 \} \cup \{ \varsigma' \mathrel{|} \varsigma \in S \text{ and } \varsigma \machstep \varsigma' \}\text. 
\end{equation*}
Then, the evaluator for the analysis computes the least fixed-point of the global transfer function:
\begin{equation*}
 \eval(e) = \mathrm{lfp}(F_{\varsigma_0})\text{,}
\end{equation*}
where $\varsigma_0 = \ev[^\mtcntr]{e, \varnothing, \varnothing, \kmt}$.

{\bf FIXME} use $\traces$

@@

A common technique to accelerate convergence in flow analyses is to share a
common, global store.
%
To retain soundness, this store grows monotonically.
%
Formally, we can cast this optimization as a second abstraction or as the
application of a widening operator during the fixed-point iteration.
%
In the ISWIM language, such a widening makes 0-CFA quartic in the size of the
program.
%
Thus, in one step, complexity drops from intractable exponentially to a merely
daunting polynomial.

@@

The only analyzers we were able to find that proved capable of
analyzing the full suite of benchmarks considered here were the Soft
Typing system of Wright and
Cartwright~\cite{dvanhorn:Wright1997Practical} and, in many ways its
successor, the Polymorphic splitting system of Wright and
Jagannathan~\cite{dvanhorn:wright-jagannathan-toplas98}.\footnote{This
  is not a coincidence; these papers set a high standard for
  evaluation, which we consciously aimed to approach.}  Unfortunately,
these analyses compute an inherently different and incomparable form
of analysis.  Consequently, we have omitted a complete comparison with
these implementations,.  The AAM approach provides more precision in
terms of temporal-ordering of program states, which comes at a cost
that can be avoided in constraint-based approaches.  Consequently
implementation techniques cannot be ``ported'' between these two
approaches.  However, our optimized implementation is within an order
of magnitude of the performance of Wright and Jaganathan's analyzer.
Although we would like to improve this to be more competitive, the
optimized AAM approach still has many strengths to recommend it in
terms of precision, ease of implementation and verification, and rapid
design. We can get closer to their performance by relying on the
representation of addresses and the behavior of $\alloc$ to
pre-allocate most data structures and split the abstract store out
into parts that are more quickly accessed and updated. The first two
of our optimizations can still be applied to an analysis that does
abstract garbage collection~\cite{dvanhorn:Might:2006:GammaCFA},
whereas the polymorphic splitting implementation is tied strongly to a
single-threaded store.

@@

Boucher and Feeley \cite{dvanhorn:Boucher1996Abstract} introduced the
idea of \emph{abstract compilation}, which used closure generation
\cite{dvanhorn:Feeley1987Using} to improve the performance of control
flow analysis.  We have adapted the closure generation technique from
composition evaluators to abstract machines and applied it to similar
effect.
